Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-04 linear_adia.tex

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth of the VSI. In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave
frequency $\hat{\omega}_c$, \emph{assuming} marginal stability to the
VSI:  
\begin{subequations}\begin{align}
    \left(\hat{\omega}_c\khat\right)^4  =& \left(\hat{\omega}_c\khat\right)^2 
  + M(M+1)\left(1-h^2q^2\khat^2\right) ,\label{relax_cond_simp1}  \\
  \left(\hat{\omega}_c\khat\right)^2 = & \frac{\beta_c}{h q} (\gamma-1)(1+2M)^2
  \left(\hat{\omega}_c\khat\right) \label{relax_cond_simp2}\\
  &- 2M(M+1) . \notag
\end{align}\end{subequations}
These equations consider $\khat^2\gg 1$, and assumes 
$|\hat{\omega}\khat| = O(1)$, $M\lesssim O(1)$ and $\beta_c\ll 1$.

Eq. \ref{relax_cond_simp1} admits the following   
approximate solution for the frequency 
if $\mathrm{min}(1,M)\times|hq\khat|^2\lesssim 1$, 
%if $4M(M+1)|hq\khat|^2/(2M+1)^2 < 1$,  
\begin{align}
  % \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left[1  + O( M|h q \khat|^2)\right]  \label{marg_freq_approx}
  \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}  
(Equality holds for $M=0$.)  
% which holds for either (a) $M= 0$ and arbitrary $|h q \hat{k}|$ or (b) $|h q \hat{k}| \ll 1$ and arbirtary $M \sim O(1)$. 
% These distinct conditions mean that our results are most robust for the fundamental vertical mode ($M = 0$) but also
% hold for higher order vertical modes when radial wavelengths are not
% too short. 
Note that Eq. \ref{marg_freq_approx} reduces to the dispersion
relation for low-frequency inertial waves in the absence of vertical
shear (see Appendix \ref{stable_novshear} for details).

If we further assume $|hq\khat|^2\ll 1$, we obtain from
Eq. \ref{relax_cond_simp2}---\ref{marg_freq_approx} that  
\begin{align}\label{bcrit_gen} 
  \beta_c &\simeq  \frac{h|q|}{(\gamma - 1)}\frac{\sqrt{1+M}}{(1+2M)}.   
%  \frac{\sqrt{1+M}}{(1+2M)}
%  \left[1 + \frac{M(2M-1)|h q   \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}%\notag
%>>>>>>> 8816838e4a06974fc61d74b84a2474adaeee3ebc
\end{align}
Eq. \ref{bcrit_gen} indicates that higher $M$ modes have lower $\beta_c$,
i.e.\ require faster cooling for VSI growth\footnote{
In fact, this conclusion does not require $\theta\equiv|hq\khat|^2$ to be extremely 
small. By retaining terms proportional to
  $\theta$, we find $\beta_c\to \beta_c \times 
  \left[1 + M(2M-1)\theta/2(2M+1)^2 \right]$. Since $\theta\lesssim 1$ by assumption, 
 $\mathrm{max}(\beta_c)$ still occurs at $M=0$.}.     
This trend is confirmed numerically in \S\ref{therm_relax_eff}.

%(and the trend holds even when the restrictive assumption
%on $|h q \hat{k}|$ is dropped).

We are most interested in the longest cooling time which allows growth for any $M$.  
This criterion is clearly set by the fundamental, $M = 0$, mode.  Our
cooling criterion for  
VSI growth is thus  
 \begin{align}\label{iso_vsi_cond}
  \beta_c\leq \beta_\mathrm{crit}  \equiv
  \frac{h|q|}{\gamma-1} . 
\end{align}
The thin disk approximation, $h \ll 1$, indicates that $\beta_c \ll
1$, as assumed to obtain Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}. 
We heuristically explain $\beta_\mathrm{crit}$ in \S\ref{vsi_require}, and numerically
test its validity in \S\ref{bcrit_num_test}.

In Appendix \ref{bcrit_alt}, we give an alternative derivation for 
$\beta_\mathrm{crit}$ which does not assume it is much less
than unity from the onset. This derivation also independent of $|h q
\khat|$.

>>>>

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth of the VSI. In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave
frequency $\hat{\omega}_c$, \emph{assuming} marginal stability to the
VSI:  
\begin{subequations}\begin{align}
    \left(\hat{\omega}_c\khat\right)^4  =& \left(\hat{\omega}_c\khat\right)^2 
  + M(M+1)\left(1-h^2q^2\khat^2\right) ,\label{relax_cond_simp1}  \\
  \left(\hat{\omega}_c\khat\right)^2 = & \frac{\beta_c}{h q} (\gamma-1)(1+2M)^2
  \left(\hat{\omega}_c\khat\right) \label{relax_cond_simp2}\\
  &- 2M(M+1) . \notag
\end{align}\end{subequations}
These equations consider $\khat^2\gg 1$, $M\lesssim O(1)$; and assume 
$|\hat{\omega}\khat| = O(1)$ and $\beta_c\ll 1$.   Note that
Eq. \ref{relax_cond_simp1} reduces to the
dispersion relation for low-frequency inertial waves in the absence of
vertical shear (see Appendix \ref{stable_novshear} for details).

In Appendix \ref{max_cool}, we find that %if $|hq\khat|^2\leq 1$ then 
\begin{align}
  |hq\khat|^2 \leq 1 \Rightarrow \frac{d\beta_c}{dM} < 0 \quad \forall
  M\geq 0. \label{bcrit_gen}
\end{align}
Eq. \ref{bcrit_gen} indicates that higher $M$ modes have lower $\beta_c$,
i.e.\ require faster cooling for VSI growth. This trend is confirmed
numerically in \S\ref{therm_relax_eff}.

% Eq. \ref{relax_cond_simp1} admits the following   
% approximate solution for the frequency 
% if $\mathrm{min}(1,M)\times|hq\khat|^2\lesssim 1$, 
% \begin{align}
%   % \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left[1  + O( M|h q \khat|^2)\right]  \label{marg_freq_approx}
%   \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left(1 -
%     \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
% \end{align}  
% (Equality holds for $M=0$.)  
% Note that Eq. \ref{marg_freq_approx} reduces to the dispersion
% relation for low-frequency inertial waves in the absence of vertical
% shear (see Appendix \ref{stable_novshear} for details). 
% If we further assume $|hq\khat|^2\ll 1$, we obtain from
% Eq. \ref{relax_cond_simp2}---\ref{marg_freq_approx} that  
% \begin{align}\label{bcrit_gen} 
%   \beta_c &\simeq  \frac{h|q|}{(\gamma - 1)}\frac{\sqrt{1+M}}{(1+2M)}.   
% %  \frac{\sqrt{1+M}}{(1+2M)}
% %  \left[1 + \frac{M(2M-1)|h q   \khat|^2}{2(2M+1)^2}\right].
% %  &\lesssim \beta_\mathrm{crit}%\notag
% %>>>>>>> 8816838e4a06974fc61d74b84a2474adaeee3ebc
% \end{align}
% Eq. \ref{bcrit_gen} indicates that higher $M$ modes have lower $\beta_c$,
% i.e.\ require faster cooling for VSI growth. 
%\footnote{
%In fact, this conclusion does not require $\theta\equiv|hq\khat|^2$ to be extremely 
%small. By retaining terms proportional to
% $\theta$, we find $\beta_c\to \beta_c \times 
%  \left[1 + M(2M-1)\theta/2(2M+1)^2 \right]$. Since $\theta\lesssim 1$ by assumption, 
% $\mathrm{max}(\beta_c)$ still occurs at $M=0$.}.     
%(and the trend holds even when the restrictive assumption
%on $|h q \hat{k}|$ is dropped).

We are most interested in the longest cooling time which allows growth for any $M$.  
This criterion is clearly set by the fundamental mode, $M = 0$. It is straight
forward to solve Eq. \ref{relax_cond_simp1} --- \ref{relax_cond_simp2}
for $\beta_c(M=0)  \equiv \beta_\mathrm{crit}$ in this special case. We thus find the cooling criterion for  
VSI growth is  
 \begin{align}\label{iso_vsi_cond}
   \beta <   \beta_\mathrm{crit}  \equiv
   \frac{h|q|}{\gamma-1} . 
 \end{align}
 The thin disk approximation, $h \ll 1$, indicates that $\beta_\mathrm{crit} \ll
 1$, as assumed to obtain Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}. 
 We heuristically explain $\beta_\mathrm{crit}$ in \S\ref{vsi_require}, and numerically
 test its validity in \S\ref{bcrit_num_test}.

We comment that the expression for $\beta_\mathrm{crit}$ may in fact 
be obtained without assuming it is much less than unity from the onset or 
restrictions on $| h q \khat|$.  This is clarified in Appendix
\ref{bcrit_alt}.
